\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

This text was assembled using various pages from
\href{https://www.wikipedia.org}{Wikipedia} as my source. I copied text
passages from many Wikipedia pages, remixed, edited, adapted and
suplemented it with my own words to assemble the chapters of this text.
I make this text available as an open educational resource under
\href{https://creativecommons.org/licenses/by-sa/3.0/deed.en}{Creative
Commons Attribution-Share Alike 3.0 Unported} United States License
(\href{https://en.wikipedia.org/wiki/Wikipedia:Text_of_Creative_Commons_Attribution-ShareAlike_3.0_Unported_License}{same}
as Wikipedia) for others to do as I did and freely copy, remix, edit and
adapt as required. Illustrations with associated hyperlinks are from
Wikimedia Commons, which should be consulted for copyright details.
Pictures and illustrations that I generated myself are subject to the
\href{https://creativecommons.org/licenses/by-sa/3.0/deed.en}{Creative
Commons Attribution-Share Alike 3.0 Unported} United States License. I
want to thank all the contributors and supporters of Wikipedia for
making it possible and keeping it relevant. I am grateful for the
support provided by the
\href{http://libguides.rcc.mass.edu/OER/proposal}{Roxbury Community
College's OER initiative} ``RCC Go Open'' and wish to thank the members
of the OER Task Force and library director William Hoag for their
unwavering commitment to OER. With your support and help this text can
only get better over time. I urge you to email me with your criticisms
and suggestions at nsucher@rcc.mass.edu.
